Follow-up Samples

### 2017-01-15 submission/manuscript_outline_20161024.Rmd
complete discussion additions

Interestingly, when previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.  This data suggests that what may change after surgery is not necessarily any one specific bacterium nor any addition of new or depressed bacterium but rather that  a shift of the bacterial community as a whole occurs with the existing members of this microbiome.

@@

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool (**insert citation**).  Based on previous work by Jobin, et al. (**insert citation**)  it may not be surprising to see E.coli in the top 5 OTUs for this model.  Similarily, Porphyromonas has also been implicated in colorectal cancer (**insert citation**).  Interestingly, many of the other OTUs had taxonomic identification for resident gut microbes.  This could suggest that changes to the resident microbiome are important to the initiation of adenoma or carcinoma formation (**insert citation**) and provide support for the hypothesis that an initial change in the bacterial microbiome could pave the way for more inflammatory species: whether by creation of a new niche for oral microbes (**insert citation**) or allowing for a bloom of existing pro-inflammatory residents (**insert citation**).

Naturally, it is curious that normal staples of many screening studies such as Fusobacterium, Parvimonas, and Peptostreptococcus were not present in the majority of the training models.  One potential explanation for this is that FIT provides the same information to the model as these three organisms and so the model uses FIT preferentially over them.  This has been suggested to be the case in a previous study (**insert Baxter Study**).  It is also possible that these specific bacteria play a major role in the progression to carcinoma but may not be as important in the initiation of an adenoma, which would be supported by our data [Figure 5].  Regardless, our study does not argue against the importance of these bacterium in CRC initiation or pathogenesis but rather that the model does not utilize these specific bacteria for prediction purposes.  Another potential reason why we did not identify the "usual suspects"" is that these bacteria may not change much between initial and follow up samples in those with an identified lesion.  That is to say that the bacteria are consistently present even after removal of the lesion by surgery.  Finally, it is likely that within our test set there was not enough indviduals in which detection  was made or relative abundance high enough for these bacteria to be significant using a paired wilcoxson test.

One limitation in this study is that we do not know whether individuals in our test set eventually had a subsequent CRC diagnosis.  This information would help to strengthen the case for our Random Forest based model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.  Another limitation is that we do not know if adding modern tests such as the stool DNA test (**insert citation**) could help improve our overall AUC.  Another limitation is that this study drew heavily from those with caucasian ancestry.  The results may not be immediately representative of those with either Asian or African ancestry.  Finally, although our training and test set are relatively large we still run the risk of overfitting or having a model that may not be immediately extrapolateable to other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.

By adding patient data such as age, BMI, etc. to the model and showing that it can successfully help to predict both carcinoma and adenoma our study provides further data that these patient factors in conjunction with the bacterial microbiome could potentially influence CRC and perhaps have a role in formation of adenomas.  Further studies need to be carried out to verify our findings since not only are we dealing with stool, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.

Despite these limitations we think that these findings significantly add to the existing scientific knowledge on CRC and the bacterial microbiome.  The ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures associated with these lesions.  It also shows that these algorithms can not only successfully react to successful treatment regimens but also may be able to one day diagnose CRC with a high level of accuracy.

>>>>

Interestingly, when only previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.  This data suggests that what may change after surgery is not necessarily any one specific bacterium nor any addition of new or depressed bacterium but rather that  a shift of the bacterial community as a whole occurs with the existing members of this microbiome.

@@

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014] by exploring what happens to the bacterial community after surgical removal of a lesion.  Based on previous work by Arthur, et al. [@arthur_intestinal_2012]  it may not be surprising to see E.coli as one of the most important OTUs and one that was common to both models.  Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.  This could suggest that the bacterial community is one of the first components that could change during the pathogenesis of disease.  These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  One potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.  An alternative reason would be that even if they were reduced in relative abundance they were still present.  Another potential explanation is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Finally, it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.  It is also possible that all three of these potential explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery.  It should be stated that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or initial and follow up).  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].

One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.  This information would help to strengthen the case for our Random Forest based model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.  Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.  Another limitation is that this study drew heavily from those with caucasian ancestry.  The results may not be immediately representative of those with either Asian or African ancestry.  Finally, although our training and test set are relatively large we still run the risk of overfitting or having a model that may not be immediately extrapolateable to other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.

Within figure 3 the initial and follow up model showed better test AUC results then the training set AUC.  This may have occured because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be overfit since the total number of samples is not that large.  However, the lesion model did not suffer from these discrepencies and similar conclusions can be drawn solely from this model.  Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with stool, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.

Despite these limitations we think that these findings significantly add to the existing scientific knowledge on CRC and the bacterial microbiome.  The ability for machine learning algorithms to take bacterial microbiome data and successfully lower positive probability after either adenoma or carcinoma removal provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with these lesions.  It also shows that these algorithms can not only successfully react to successful treatment regimens but also may be able to one day stratify CRC disease risk with a high level of accuracy.
